---
title: "CBASS Data Cleaning and Exploration"
author: "Katie Lankowicz"
date: "2023-06-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning= FALSE)
```

```{r libraries}
# Clear workspace
rm(list=ls())

# Load packages
library(reprex, quietly=T, verbose=F)
suppressPackageStartupMessages(library(tidyverse, quietly=T, verbose=F))
suppressPackageStartupMessages(library(readxl, quietly=T, verbose=F))
suppressPackageStartupMessages(library(here, quietly=T, verbose=F))
suppressPackageStartupMessages(library(padr, quietly=T, verbose=F))
suppressPackageStartupMessages(library(zoo, quietly=T, verbose=F))
suppressPackageStartupMessages(library(sf, quietly=T, verbose=F))
suppressPackageStartupMessages(library(vegan, quietly=T, verbose=F))
#suppressPackageStartupMessages(library(kableExtra, quietly=T, verbose=F))
#suppressPackageStartupMessages(library(reshape2, quietly=T, verbose=F))
#suppressPackageStartupMessages(library(ggsflabel, quietly=T, verbose=F))
#suppressPackageStartupMessages(library(cowplot, quietly=T, verbose=F))
#suppressPackageStartupMessages(library(smplot2, quietly=T, verbose=F))

# Negate function
'%notin%' <- function(x,y)!('%in%'(x,y))

# Set GGplot auto theme
theme_set(theme(panel.grid.major = element_line(color='lightgray'),
                panel.grid.minor = element_blank(),
                panel.background = element_blank(),
                panel.border = element_rect(color='black', linewidth=1, fill=NA),
                legend.position = "bottom",
                axis.text.x=element_text(size=12),
                axis.text.y=element_text(size=12),
                axis.title.x=element_text(size=14),
                axis.title.y=element_text(size=14, angle=90, vjust=2),
                plot.title=element_text(size=14, hjust = 0, vjust = 1.2),
                plot.caption=element_text(hjust=0, face='italic', size=12),
                strip.text.x=element_text(size=12)))
```

## Casco Bay Aquatic Systems Survey
The CBASS project has been ongoing since 2014, with a seine component in the greater Portland region of Casco Bay every year. The purpose of this document is to create a protocol for data cleaning and combination for further analysis, then do some basic data exploration.

The data for 2014 - 2021 (excluding 2019) are stored in the Excel workbook `raw-seine-data.xlsx` across the sheets named `sites`, `species`, `trips`, and `fish`. This sheet naming convention is kept for all Excel workbooks containing seine data. The data for 2022 are stored in the Excel workbook `2022_Raw_Seine_Data.xlsx`. The data for the latter half of 2018 and a few trips in 2019 are stored in `Missing_2018_2019_Seine_Data.xlsx` and were entered in 2023. There are only 2 trips noted for 2019, which may be fewer than actually took place. The data for 2023 are currently being entered and will be included in a later update.

### Data loading and combination
Data from Excel workbooks will be loaded in such a way that the individual sheets within the workbook will become dataframes within a list item in R. We will keep only the sheets mentioned above in cases where other data are provided. It's also important to ensure that variable names (column names) are standardized so that we can merge data from all years into a single dataframe later on.

#### Load and clean 2014-2021 data
The bulk of the data are stored in this Excel sheet. There are extra sheets to remove, quality control issues to address, and site names to standardize. The first step is to load the data as a list of dataframes, one dataframe for each Excel sheet in the Excel workbook. Necessary dataframes will be saved, and extraneous (blank) columns in these dataframes will be removed.

```{r loading2014, results='hide', message=F, warning=F}
# 2014-2021 data
  # Set the name of the workbook
  fname <- paste0(here(), '/Raw_Data/raw-seine-data_061621.xlsx')
  # Get info about all excel sheet names in workbook
  sheets <- readxl::excel_sheets(fname)
  # Read in as list item, each item is a sheet
  tibble <- invisible(lapply(sheets, 
                             function(x) readxl::read_excel(fname, sheet = x)))
  # Coerce list items to dataframes
  data_frame <- lapply(tibble, as.data.frame)
  # Name list items
  names(data_frame) <- sheets
  # Create temporary dataframe with all sheets
  data.2014.temp <- data_frame
  # Create new, empty data list
  data.2014 <- vector(mode="list", length = 5)
  # Name list items
  names(data.2014) <- c('sites', 'species', 'trips', 'fish', 'abundance')
  # Save pertinent information in matching list items
  data.2014$sites <- data.2014.temp$sites
  data.2014$species <- data.2014.temp$species
  data.2014$trips <- data.2014.temp$trips
  data.2014$fish <- data.2014.temp$fish
  data.2014$abundance <- data.2014.temp$abundance
  #data.2014$abundance$notes <- data.2014$abundance$...6
  # Remove intermediates
  rm(data.2014.temp, data_frame, tibble, fname, sheets)
  
  # Fix column names in each list item
  colnames(data.2014$sites) <- c('site_id', 'site_name', 'latitude', 'longitude',
                                 'utm_easting', 'utm_northing')
  colnames(data.2014$species) <- c('ID', 'common_name', 'nhfg_code')
  data.2014$trips <- dplyr::select(data.2014$trips,
                                   trip_id, date, site_id, site_name, substrate,
                                   gear, set, set_time, weather, tide, 
                                   time_of_high_tide, ft_1, 
                                   time_of_low_tide, ft_2,
                                   temp_degc, do_percent, `do_mg/l`, conductivity,
                                   salinity_ppt, hermit_crabs, shrimp,
                                   macroalgae, notes)
  colnames(data.2014$fish) <- c('fish_id', 'trip_id', 'sample_id', 'species_name',
                                'species_code', 'sex', 'field_length_mm', 
                                'lab_length_mm', 'weight_g', 'notes')
  data.2014$abundance <- dplyr::select(data.2014$abundance,
                                       trip_id, species_name,
                                       catch)

```

The sites dataframe needs to be cleaned to ensure sites are referred to in the same way across the years of the dataset. We will also remove freshwater sites at Highland Lake and any sites that were sampled fewer than 10 times across the 8 years of the survey.

```{r siteclean2014}
#### Site Identification ####
# Remove freshwater sites and Willard Beach - WB only sampled once
data.2014$sites <- subset(data.2014$sites,
                          site_name %notin% c('Highland Lake Association Boat Ramp',
                                              'Public Boat Ramp at Highland Lake',
                                              'Cottage Road at Highland Lake',
                                              'Family House - Highland',
                                              'Willard Beach'))
# Remove The Brothers - South (only sampled 4 times across 8 years)
data.2014$sites <- data.2014$sites[data.2014$sites$site_name != "The Brothers - South",]

# Rename Gilsland Farm to Audubon and correct spelling of Skitterygusset
data.2014$sites$site_name[data.2014$sites$site_name == "Gilsland Farm"] <- 'Audubon'
data.2014$sites$site_name[data.2014$sites$site_name == 'Scitterygusset'] <- 
  'Skitterygusset'

# Assign bay locations
bay_locations <- data.frame(
  site_name = c('Presumpscot Moorings', 'Skitterygusset', 'Audubon',
                'Mussel Cove', 'The Brothers - North', 'Mackworth Island - North',
                               'Mackworth Island - Beach', 'Back Cove',
                'Great Diamond Island', 'SMCC', 'Cushing Island', 'Alewife Cove'),
  bay_location = c(rep('Inner', 3),
                   rep('Mid', 5),
                   rep('Outer', 4))
)
data.2014$sites <- left_join(bay_locations, data.2014$sites, 
                             by=c('site_name'))

# Assign new site_id numbers
data.2014$sites$site_id <- NULL
data.2014$sites$site_id <- seq(1:nrow(data.2014$sites))

# Order
data.2014$sites <- data.2014$sites[with(data.2014$sites,
                                        order(site_id)),]
rownames(data.2014$sites) <- NULL
data.2014$sites$site_id <- str_pad(data.2014$sites$site_id, 2, "left", "0")
```

The trip information also needs to be cleaned. Any trips that occurred at our removed sites need to be removed from the trip info dataframe. Variables will also be checked for validity. For example, one temperature is recorded as 147 degrees C. This clearly is a typo, and will be replaced with 14.7 degrees C. Categorical variables will be forced to set levels; there are instances in which people wrote editorialized versions of what they were supposed to, and we do not need these extra details for our quantitative analysis.

```{r tripclean2014}
#### Trip Identification #####
# Strip trip info from data list
trips <- data.2014$trips

# Remove blank rows
trips <- trips[!is.na(trips$date),]

# Remove freshwater and Willard Beach sites
trips <- trips[trips$site_name %notin% c('Cottage Road at Highland',
                                         'Cottage Road at Highland Lake',
                                         'Family House - Highland',
                                         'Highland Lake Association Boat Ramp',
                                         'Public Boat Ramp at Highland Lake',
                                         'Willard Beach',
                                         'The Brothers - South'),]
# Fix some spelling errors
trips$site_name[trips$site_name == "Cushing"] <- "Cushing Island"
trips$site_name[trips$site_name == "Gilsland Farm"] <- "Audubon"
trips$site_name[trips$site_name == "Mackworth Island-Beach"] <- 
  "Mackworth Island - Beach"
trips$site_name[trips$site_name == "Mackworth Island-North"] <-
  "Mackworth Island - North"
trips$site_name[trips$site_name == "Scitterygusset"] <- "Skitterygusset"
trips$site_name <- as.factor(trips$site_name)

# Re-assign site_id
trips$site_id <- NULL
trips <- left_join(trips, dplyr::select(data.2014$sites, site_name, site_id),
                   by=c('site_name'))

# Fix substrate data type
trips$substrate <- as.factor(trips$substrate)

# Remove instances where seine was not set
trips <- subset(trips, set == 'yes')

# Remove gear and set columns, no longer needed
trips <- dplyr::select(trips, -gear, -set)

# Fix set time
trips$set_time <- substr(trips$set_time, start = 12, stop=19)
trips$set_time <- paste0(trips$date, " ", trips$set_time)
trips$set_time <- as.POSIXct(trips$set_time,
                               format = "%Y-%m-%d %H:%M:%S")

# Remove tide information, will need to look it up and add it back.
trips <- dplyr::select(trips, -time_of_high_tide, -time_of_low_tide,
                       -ft_1, -ft_2, -tide)

# Fix weather info
trips$weather[trips$weather %in% c('calm/cloudy/70',
                                   'calm/cloudy/74',
                                   'cloudy',
                                   'overcast')] <- 'overcast'

trips$weather[trips$weather %in% c('calm/partlycloudy/70',
                                   'partly cloudy',
                                   'partly sunny',
                                   'sun/clouds',
                                   'sunny, slightly overcast')] <- 'partly cloudy'

trips$weather[trips$weather %in% c('calm/sunny',
                                   'calm/sunny/75',
                                   'hot, sunny',
                                   'mild, breezy',
                                   'sun',
                                   'sunny, calm',
                                   'sunny, east wind',
                                   'sunny/70',
                                   'sunny/calm',
                                   'sunny/little windy',
                                   'sunny/warm',
                                   'sunny/warm/calm',
                                   'sunny/windy',
                                   'sunny/windy (north wind = exposure)',
                                   'suny')] <- 'sunny'

trips$weather[trips$weather %in% c('cloudy, stormy',
                                   'rain',
                                   'rain/fog')] <- 'rain'

trips$weather[trips$weather %in% c('fog',
                                   'foggy, partly cloudy',
                                   'sun, fog')] <- 'foggy'

trips$weather <- as.factor(trips$weather)

# Remove conductivity, macroalgae
trips <- dplyr::select(trips, -conductivity, -macroalgae)

# Fix hermit crabs 
trips$hermit_crabs[trips$hermit_crabs == 2] <- 1
trips$hermit_crabs <- as.numeric(trips$hermit_crabs)

# Fix shrimp
trips$shrimp <- as.numeric(trips$shrimp)

# Create year column
trips$year <- substr(trips$date, start=1, stop = 4)
trips$year <- as.numeric(trips$year)

# Create new trip_id based on date
trips$new_trip_id <- as.numeric(as.factor(trips$date))

# Create new overall information identification
trips$loc_id <- paste0(trips$year, "_",
                                 str_pad(trips$new_trip_id, 3, "left", "0"),
                                 "_",
                                 str_pad(trips$site_id, 2, "left", "0"))

# Assign bay location
trips <- left_join(trips, bay_locations, by=c('site_name'))
trips$bay_location <- as.factor(trips$bay_location)
trips$site_name <- as.factor(trips$site_name)

# Quality control
trips$temp_degc[trips$temp_degc == 147] <- 14.7
trips$`do_mg/l`[trips$`do_mg/l` == 145] <- 14.5

# Reorder
trips <- dplyr::select(trips, loc_id, year, date, trip_id, new_trip_id, 
                       site_id, site_name, bay_location,
                       substrate, set_time, weather, temp_degc, 
                       `do_mg/l`, salinity_ppt, hermit_crabs, shrimp,
                       notes)
# Save
trips <- trips[!is.na(trips$loc_id),]
data.2014$trips <- trips
rm(trips)
```

Moving on, the biological data for the subsampled 25 individuals per species need to be cleaned. Species names will be checked for spelling errors. The information will then be linked to the new trip_id structure so fish can be assigned to a physical location and time.

```{r bioclean2014}
#### Fish Biological Information ####
bio <- data.2014$fish

# Remove anything that doesn't have a trip_id in the trips dataframe
bio <- bio[bio$trip_id %in% data.2014$trips$trip_id,]
bio <- bio[!is.na(bio$trip_id),]

# Coerce trip id to character
bio$trip_id <- as.character(bio$trip_id)

# Remove unnecessary columns
bio <- dplyr::select(bio, trip_id, species_name, sex, 
                     field_length_mm, lab_length_mm, notes)

# Assign site_name and location info
bio <- left_join(bio, dplyr::select(data.2014$trips, trip_id, new_trip_id,
                                    loc_id, site_name, site_id, bay_location,
                                    date),
                 by=c('trip_id'))

# Re-order
bio <- dplyr::select(bio, loc_id, date, trip_id, new_trip_id, site_name, site_id,
                     bay_location, species_name, sex, field_length_mm, 
                     lab_length_mm, notes)

# Fix data types
bio$site_name <- as.factor(bio$site_name)
bio$bay_location <- as.factor(bio$bay_location)

# Fix species names
bio$species_name[bio$species_name == "blue fish"] <- 'bluefish'
bio$species_name[bio$species_name == "blueback"] <- 'blueback herring'
bio$species_name[bio$species_name == "grubby scuplin"] <- 'grubby sculpin'
bio$species_name[bio$species_name == "pogie"] <- 'atlantic menhaden'
bio$species_name[bio$species_name == "sand lance"] <- 'sandlance'
bio$species_name[bio$species_name == "shad"] <- 'american shad'
bio$species_name[bio$species_name == "silverside"] <- 'atlantic silverside'
bio$species_name[bio$species_name == "tom cod"] <- 'tomcod'

bio$species_name[bio$species_name == "shiner"] <- 'unID shiner'
bio$species_name[bio$species_name == "sculpin"] <- 'unID sculpin'
bio$species_name[bio$species_name == 'sturgeon'] <- 'unID sturgeon'
bio$species_name[bio$species_name == "stickleback"] <- 'unID stickleback'
bio$species_name[bio$species_name == "unIDgunnel"] <- 'unID gunnel'
bio$species_name <- as.factor(bio$species_name)

# Remove instances of no catch
bio <- bio[bio$species_name %notin% c('no fish', 'no catch'),]

# Fix sex
bio$sex <- as.factor(bio$sex)

# Remove information on fish with no lengths
for(i in 1:nrow(bio)){
  if(is.na(bio$field_length_mm[i]) & is.na(bio$lab_length_mm[i])){
    bio$notes[i] <- 'Sample lost'
  }
}
bio$notes <- as.factor(bio$notes)
bio <- bio[bio$notes %notin% c('Sample lost'),]

# Finalize length information - if both lab and field measurements exist, use lab
for(i in 1:nrow(bio)){
  if(!is.na(bio$field_length_mm[i]) & is.na(bio$lab_length_mm[i])){
    bio$length_mm[i] <- bio$field_length_mm[i]
  }
  if(!is.na(bio$lab_length_mm[i]) & is.na(bio$field_length_mm[i])){
    bio$length_mm[i] <- bio$lab_length_mm[i]
  }
  if(!is.na(bio$field_length_mm[i]) & !is.na(bio$lab_length_mm[i])){
    bio$lab_length_mm[i] <- bio$lab_length_mm[i]
  }
}

# There is definitely not a 1mm horseshoe crab or periwinkle
bio <- subset(bio, bio$length_mm != 1)

# Quality control
bio <- bio[!is.na(bio$species_name),]

# Finalize dataframe
bio <- dplyr::select(bio,
                     loc_id, date, site_name, site_id, bay_location, 
                     species_name, sex, length_mm, notes)

data.2014$fish <- bio
rm(bio)
```

The abundance data need to be cleaned in a similar manner to the biological data.

```{r abundclean2014}
#### Abundance information ####
abund <- data.2014$abundance

# Coerce trip id to character
abund$trip_id <- as.character(abund$trip_id)

# Remove anything that doesn't have a trip_id in the trips dataframe
abund <- abund[abund$trip_id %in% data.2014$trips$trip_id,]
abund <- abund[!is.na(abund$trip_id),]

# Assign site_name and location info
abund <- left_join(abund, dplyr::select(data.2014$trips, trip_id, new_trip_id,
                                    loc_id, site_name, site_id, bay_location,
                                    date),
                 by=c('trip_id'))

# Re-order
abund <- dplyr::select(abund, loc_id, date, trip_id, new_trip_id, site_name, 
                       site_id, bay_location, 
                       species_name, catch)
abund <- abund[complete.cases(abund),]

# Fix data types
abund$site_name <- as.factor(abund$site_name)
abund$bay_location <- as.factor(abund$bay_location)

# Coerce non-numerics to numerics
abund$catch2 <- abund$catch
abund$catch[abund$catch == "~50"] <- 50
abund$catch[abund$catch == "20+"] <- 20
abund$catch[abund$catch == "25+"] <- 25
abund$catch[abund$catch == "6+"] <- 6
abund$catch[abund$catch == "9+"] <- 9

# Remove 0 catch instances
abund <- abund[abund$catch != '0',]

# Remove useless information
abund <- abund[abund$catch %notin% c('lots', 'x'),]
abund$catch <- as.numeric(abund$catch)
abund$catch2 <- NULL

# Fix species names
abund <- abund[abund$species_name %notin% c('?', 
                                            'cusk? hake?',
                                            'unid'),]

abund$species_name[abund$species_name == "blue fish"] <- 'bluefish'
abund$species_name[abund$species_name == "blueback"] <- 'blueback herring'
#abund$species_name[abund$species_name == "grubby scuplin"] <- 'grubby sculpin'
abund$species_name[abund$species_name == "pogie"] <- 'atlantic menhaden'
abund$species_name[abund$species_name == "pogy"] <- 'atlantic menhaden'
abund$species_name[abund$species_name == "sand lance"] <- 'sandlance'
abund$species_name[abund$species_name == "shad"] <- 'american shad'
abund$species_name[abund$species_name == "silverside"] <- 'atlantic silverside'
#abund$species_name[abund$species_name == "tom cod"] <- 'tomcod'

abund$species_name[abund$species_name == "shiner"] <- 'unID shiner'
abund$species_name[abund$species_name == "sculpin"] <- 'unID sculpin'
abund$species_name[abund$species_name == 'sturgeon'] <- 'unID sturgeon'
abund$species_name[abund$species_name == "stickleback"] <- 'unID stickleback'
abund$species_name[abund$species_name == "unIDgunnel"] <- 'unID gunnel'
abund$species_name <- as.factor(abund$species_name)

# Finalize
abund <- dplyr::select(abund, -trip_id, -new_trip_id)
data.2014$abundance <- abund

```

Finally, the data will be QA-QC'd to ensure all variables are the correct format. We will also ensure that the number of fish reported in the abundance dataframe makes sense given the number of fish in the biological information dataframe. Recall that only 25 fish from each species are subsampled for biological information at each site; frequently, we will report higher abundance than number of fish measured However, we should never have more fish measured than reported in the abundance information. If this happens, we will use the number of fish measured as the correct number caught.

```{r finalcheck2014}
#str(data.2014$sites)
data.2014$sites$site_name <- as.factor(data.2014$sites$site_name)
data.2014$sites$bay_location <- as.factor(data.2014$sites$bay_location)
data.2014$trips <- dplyr::select(data.2014$trips, -trip_id, -new_trip_id)

test.bio <- data.2014$fish
test.bio <- dplyr::select(test.bio,
                          loc_id, date, species_name, length_mm)
test.bio <- test.bio %>%
  count(loc_id, species_name)

test.abund <- data.2014$abundance
test.abund <- dplyr::select(test.abund,
                            loc_id, date, species_name, catch)

test.combo <- left_join(test.abund, test.bio, by=c("loc_id", "species_name"))

test.bad <- test.combo[test.combo$n != test.combo$catch,]
test.bad <- test.bad[test.bad$n %notin% c(24,25, 26),]
test.bad$dif <- test.bad$n - test.bad$catch

test.bad <- test.bad[with(test.bad, order(dif)),]
test.bad <- test.bad[complete.cases(test.bad),]
test.bad <- test.bad[test.bad$dif > 0,]

data.2014$abundance$catch[data.2014$abundance$loc_id == "2015_023_05" &
                            data.2014$abundance$species_name == "mummichog"] <- 7
data.2014$abundance$catch[data.2014$abundance$loc_id == "2016_033_05" &
                            data.2014$abundance$species_name == "green crab"] <- 3
data.2014$abundance$catch[data.2014$abundance$loc_id == "2016_036_03" &
                            data.2014$abundance$species_name == "green crab"] <- 2
data.2014$abundance$catch[data.2014$abundance$loc_id == "2017_042_08" &
                            data.2014$abundance$species_name == "atlantic herring"] <- 10
data.2014$abundance$catch[data.2014$abundance$loc_id == "2021_069_06" &
                            data.2014$abundance$species_name == "atlantic herring"] <- 5
data.2014$abundance$catch[data.2014$abundance$loc_id == "2016_037_10" &
                            data.2014$abundance$species_name == "green crab"] <- 4
data.2014$abundance$catch[data.2014$abundance$loc_id == "2021_074_07" &
                            data.2014$abundance$species_name == "atlantic silverside"] <- 12
data.2014$abundance$catch[data.2014$abundance$loc_id == "2021_074_07" &
                            data.2014$abundance$species_name == "green crab"] <- 14
data.2014$abundance$catch[data.2014$abundance$loc_id == "2021_076_10" &
                            data.2014$abundance$species_name == "mummichog"] <- 7
data.2014$abundance$catch[data.2014$abundance$loc_id == "2021_078_11" &
                            data.2014$abundance$species_name == "green crab"] <- 15
data.2014$abundance$catch[data.2014$abundance$loc_id == "2020_061_05" &
                            data.2014$abundance$species_name == "alewife"] <- 28
```

#### Load and clean 2022 data
Data only from 2022 are stored in this Excel sheet. There are extra sheets to remove, quality control issues to address, and site names to standardize. The first step is to load the data as a list of dataframes, one dataframe for each Excel sheet in the Excel workbook. Necessary dataframes will be saved, and extraneous (blank) columns in these dataframes will be removed.

```{r loading2022, results='hide', message=F, warning=F}
# 2022 data
  # Get info about all excel sheets in workbook
  fname <- paste0(here(), '/Raw_Data/2022_Raw_Seine_Data.xlsx')
  sheets <- readxl::excel_sheets(fname)
  # Read in as list item, each item is a sheet
  tibble <- invisible(lapply(sheets, 
                             function(x) readxl::read_excel(fname, sheet = x)))
  # Coerce list items to dataframes
  data_frame <- lapply(tibble, as.data.frame)
  # Name list items
  names(data_frame) <- sheets
  # Create temporary dataframe with all sheets
  data.2022.temp <- data_frame
  # Cut to new data list
  data.2022 <- vector(mode="list", length = 5)
  names(data.2022) <- c('sites', 'species', 'trips', 'fish', 'abundance')
  data.2022$sites <- data.2022.temp$Sites
  data.2022$species <- data.2022.temp$Species
  data.2022$trips <- data.2022.temp$Trips
  data.2022$fish <- data.2022.temp$`Fish Length`
  data.2022$abundance <- data.2022.temp$`Fish Abundance`
  # Remove intermediates
  rm(data.2022.temp, data_frame, tibble, fname, sheets)
  
    # Fix names
  colnames(data.2022$sites) <- c('site_id', 'site_name', 'latitude', 'longitude',
                                 'utm_easting', 'utm_northing')
  colnames(data.2022$species) <- c('common_name', 'nhfg_code')
  data.2022$species <- left_join(data.2022$species, data.2014$species,
                                 by=c('common_name', 'nhfg_code'))
  data.2022$trips <- dplyr::select(data.2022$trips,
                                   trip_id, date, site_id, site_name, substrate,
                                   gear, set, set_time, weather, tide, 
                                   time_of_high_tide, ft_1, 
                                   time_of_low_tide, ft_2,
                                   temp_degc, do_percent, `do_mg/l`, conductivity,
                                   salinity_ppt, hermit_crabs, shrimp,
                                   notes)
  data.2022$trips$macroalgae <- NA
  colnames(data.2022$fish) <- c('fish_id', 'trip_id', 'sample_id', 'species_name',
                                'species_code', 'sex', 'field_length_mm', 
                                'lab_length_mm', 'weight_g', 'notes')
  colnames(data.2022$abundance) <- c('trip_id', 'species_name', 'catch')

```

The site information does not need to be cleaned. We will copy the site dataframe from 2014-2021 to the 2022 dataframe, because the same sites were sampled in these two periods.

```{r siteclean2022}
#### Site Identification ####
# Sites will be the same year to year.
# What is called "Brothers" in 2022 is the same as "The Brothers - North" in 
# previous years.
data.2022$sites <- data.2014$sites
```

The trip information will be cleaned similar to the 2014-2021 dataset. Freshwater sites do not exist in this year, but there are spelling errors to address. The site numbers and loc_id variables will be merged to match the patterns that exist in the 2014-2021 dataset.

```{r tripclean2022}
#### Trip Identification #####
# Strip trip info from data list
trips <- data.2022$trips

# Remove blank rows
trips <- trips[!is.na(trips$date),]

# Fix some spelling errors
trips$site_name[trips$site_name == "Cushing"] <- "Cushing Island"
trips$site_name[trips$site_name == "Mack Beach"] <- 
  "Mackworth Island - Beach"
trips$site_name[trips$site_name == "Mack North"] <-
  "Mackworth Island - North"
trips$site_name[trips$site_name == "Scitterygusset"] <- "Skitterygusset"
trips$site_name[trips$site_name == "Presumpscot"] <- "Presumpscot Moorings"
trips$site_name[trips$site_name == "Brothers"] <- "The Brothers - North"
trips$site_name[trips$site_name == "Great Diamond"] <- "Great Diamond Island"

trips$site_name <- as.factor(trips$site_name)

# Re-assign site_id
trips$site_id <- NULL
trips <- left_join(trips, dplyr::select(data.2022$sites, site_name, site_id),
                   by=c('site_name'))

# Fix substrate data type
trips$substrate <- as.factor(trips$substrate)

# Remove instances where seine was not set
trips <- subset(trips, set == 'yes')

# Remove gear and set columns, no longer needed
trips <- dplyr::select(trips, -gear, -set)

# Fix set time
trips$set_time <- substr(trips$set_time, start = 12, stop=19)
trips$set_time <- paste0(trips$date, " ", trips$set_time)
trips$set_time <- as.POSIXct(trips$set_time,
                               format = "%Y-%m-%d %H:%M:%S")

# Remove tide information, will need to look it up and add it back.
trips <- dplyr::select(trips, -time_of_high_tide, -time_of_low_tide,
                       -ft_1, -ft_2, -tide)

# Fix weather info
trips$weather[trips$weather %in% c('cloudy',
                                   'cloudy/foggy',
                                   'low gray clouds',
                                   'overcast')] <- 'overcast'

trips$weather[trips$weather %in% c('partly cloudy',
                                   'partly cloudy/sun',
                                   'partly sunny')] <- 'partly cloudy'

trips$weather[trips$weather %in% c('mostly sunny',
                                   'mostly sunny, wind picking up')] <- 
  'mostly sunny'

trips$weather[trips$weather %in% c('rain',
                                   'rain started')] <- 'rain'

trips$weather[trips$weather %in% c('clear skies',
                                   'sunny',
                                   'sunny, very windy',
                                   'sunny/windy')] <- 'sunny'

trips$weather <- as.factor(trips$weather)

# Remove conductivity, macroalgae
trips <- dplyr::select(trips, -conductivity, -macroalgae)

# Fix hermit crabs 
trips$hermit_crabs[trips$hermit_crabs == 'present'] <- 1
trips$hermit_crabs[trips$hermit_crabs == 'absent'] <- 0
trips$hermit_crabs[trips$hermit_crabs == 'not circled'] <- NA
trips$hermit_crabs <- as.numeric(trips$hermit_crabs)

# Fix shrimp
trips$shrimp[trips$shrimp == 'absent'] <- 0
trips$shrimp[trips$shrimp == 'low'] <- 1
trips$shrimp[trips$shrimp == 'medium'] <- 2
trips$shrimp[trips$shrimp == 'high'] <- 3
trips$shrimp[trips$shrimp == 'very, very high'] <- 3
trips$shrimp[trips$shrimp == 'not circled'] <- NA
trips$shrimp <- as.numeric(trips$shrimp)

# Create year column
trips$year <- substr(trips$date, start=1, stop = 4)
trips$year <- as.numeric(trips$year)

# Create new trip_id based on date
trips$new_trip_id <- as.numeric(as.factor(trips$date)) +
  max(as.numeric(as.factor(data.2014$trips$date)))

# Create new overall information identification
trips$loc_id <- paste0(trips$year, "_",
                                 str_pad(trips$new_trip_id, 3, "left", "0"),
                                 "_",
                                 str_pad(trips$site_id, 2, "left", "0"))

# Assign bay location
trips <- left_join(trips, bay_locations, by=c('site_name'))

# Reorder
trips <- dplyr::select(trips, loc_id, year, date, trip_id, new_trip_id, 
                       site_id, site_name, bay_location,
                       substrate, set_time, weather, temp_degc, 
                       `do_mg/l`, salinity_ppt, hermit_crabs, shrimp,
                       notes)
# QA-QC
trips$bay_location <- as.factor(trips$bay_location)

# Save
trips <- trips[!is.na(trips$loc_id),]
data.2022$trips <- trips
rm(trips)
```

Biological subsample information will be cleaned, but it is generally correct and doesn't need much adjustment.

```{r bioclean2022}
#### Fish Biological Information ####
bio <- data.2022$fish

# Remove anything that doesn't have a trip_id in the trips dataframe
bio <- bio[bio$trip_id %in% data.2022$trips$trip_id,]
bio <- bio[!is.na(bio$trip_id),]

# Remove unnecessary columns
bio <- dplyr::select(bio, trip_id, species_name, sex, 
                     field_length_mm, lab_length_mm, notes)

# Assign site_name and location info
bio <- left_join(bio, dplyr::select(data.2022$trips, trip_id, new_trip_id,
                                    loc_id, site_name, site_id, bay_location,
                                    date),
                 by=c('trip_id'))

# Re-order
bio <- dplyr::select(bio, loc_id, date, trip_id, new_trip_id, site_name, site_id,
                     bay_location, species_name, sex, field_length_mm, 
                     lab_length_mm, notes)

# Fix data types
bio$site_name <- as.factor(bio$site_name)
bio$bay_location <- as.factor(bio$bay_location)

# Fix species names
bio$species_name[bio$species_name == "pipefish"] <- 'northern pipefish'
bio$species_name[bio$species_name == "sand lance"] <- 'sandlance'
bio$species_name[bio$species_name == "silverside"] <- 'atlantic silverside'

bio$species_name[bio$species_name == "sculpin"] <- 'unID sculpin'
bio$species_name[bio$species_name == 'sturgeon'] <- 'unID sturgeon'
bio$species_name <- as.factor(bio$species_name)

# Remove instances of no catch
bio <- bio[bio$species_name %notin% c('no fish', 'no catch'),]

# Fix sex
bio$sex <- as.factor(bio$sex)

# Remove information on fish with no lengths
for(i in 1:nrow(bio)){
  if(is.na(bio$field_length_mm[i]) & is.na(bio$lab_length_mm[i])){
    bio$notes[i] <- 'Sample lost'
  }
}
bio$notes <- as.factor(bio$notes)
bio <- bio[bio$notes %notin% c('Sample lost'),]

# Finalize length information - if both lab and field measurements exist, use lab
for(i in 1:nrow(bio)){
  if(!is.na(bio$field_length_mm[i]) & is.na(bio$lab_length_mm[i])){
    bio$length_mm[i] <- bio$field_length_mm[i]
  }
  if(!is.na(bio$lab_length_mm[i]) & is.na(bio$field_length_mm[i])){
    bio$length_mm[i] <- bio$lab_length_mm[i]
  }
  if(!is.na(bio$field_length_mm[i]) & !is.na(bio$lab_length_mm[i])){
    bio$lab_length_mm[i] <- bio$lab_length_mm[i]
  }
}

# Finalize dataframe
bio <- dplyr::select(bio,
                     loc_id, date, site_name, site_id, bay_location, 
                     species_name, sex, length_mm, notes)
bio <- bio[!is.na(bio$species_name),]

data.2022$fish <- bio
rm(bio)
```

The same is true of the abundance information.

```{r abundclean2022}
#### Abundance information ####
abund <- data.2022$abundance

# Remove anything that doesn't have a trip_id in the trips dataframe
abund <- abund[abund$trip_id %in% data.2022$trips$trip_id,]
abund <- abund[!is.na(abund$trip_id),]

# Assign site_name and location info
abund <- left_join(abund, dplyr::select(data.2022$trips, trip_id, new_trip_id,
                                    loc_id, site_name, site_id, bay_location,
                                    date),
                 by=c('trip_id'))

# Re-order
abund <- dplyr::select(abund, loc_id, date, trip_id, new_trip_id, site_name, 
                       site_id, bay_location, 
                       species_name, catch)
abund <- abund[complete.cases(abund),]

# Fix data types
abund$site_name <- as.factor(abund$site_name)
abund$bay_location <- as.factor(abund$bay_location)

# Coerce non-numerics to numerics
abund$catch <- as.numeric(abund$catch)

# Fix species names
abund$species_name[abund$species_name == "pipefish"] <- 'northern pipefish'
abund$species_name[abund$species_name == "sand lance"] <- 'sandlance'
abund$species_name[abund$species_name == "silverside"] <- 'atlantic silverside'
abund$species_name[abund$species_name == 'squid'] <- 'shortfin squid'

abund$species_name[abund$species_name == "sculpin"] <- 'unID sculpin'
abund$species_name[abund$species_name == 'sturgeon'] <- 'unID sturgeon'
abund$species_name <- as.factor(abund$species_name)

# Finalize
abund <- dplyr::select(abund, -trip_id, -new_trip_id)
data.2022$abundance <- abund
rm(abund)
```

```{r finalclean2022}
#str(data.2022$sites)
#str(data.2022$trips)
data.2022$trips$site_name <- as.factor(data.2022$trips$site_name)
data.2022$trips$bay_location <- as.factor(data.2022$trips$bay_location)
data.2022$trips <- dplyr::select(data.2022$trips, -trip_id, -new_trip_id)
#str(data.2022$fish)
#str(data.2022$abundance)

test.bio <- data.2022$fish
test.bio <- dplyr::select(test.bio,
                          loc_id, date, species_name, length_mm)
test.bio <- test.bio %>%
  count(loc_id, species_name)

test.abund <- data.2022$abundance
test.abund <- dplyr::select(test.abund,
                            loc_id, date, species_name, catch)

test.combo <- left_join(test.abund, test.bio, by=c("loc_id", "species_name"))

test.bad <- test.combo[test.combo$n != test.combo$catch,]
test.bad <- test.bad[test.bad$n %notin% c(24,25, 26),]
test.bad$dif <- test.bad$n - test.bad$catch

test.bad <- test.bad[with(test.bad, order(dif)),]
test.bad <- test.bad[complete.cases(test.bad),]
test.bad <- test.bad[test.bad$dif > 0,]

# No issues with catch-bio discrepancies
rm(test.abund, test.bio, test.combo, i, test.bad)
```
 
As a final check, we will ensure that species names are represented correctly between biological information and abundance datasets.

```{r abundcheck, echo=F, results='hide'}
#### Present in biological information but not in abundance ####
# 2014
as.character(data.2014$fish$species_name[data.2014$fish$species_name %notin%
                              data.2014$abundance$species_name])
date.check <- unique(data.2014$fish$loc_id[data.2014$fish$species_name == 
                                      "shortnose sculpin"])
data.2014$abundance[data.2014$abundance$loc_id == paste0(date.check),]
# So this is a typo. What is in the biological data as "shortnose" should actually
# be "shorthorn".
data.2014$fish$species_name[data.2014$fish$species_name == "shortnose sculpin"] <- 
  "shorthorn sculpin"

date.check <- unique(data.2014$fish$loc_id[data.2014$fish$species_name == 
                                      "unID stickleback"])
data.2014$abundance[data.2014$abundance$loc_id == paste0(date.check[1]),]
data.2014$abundance[data.2014$abundance$loc_id == paste0(date.check[2]),]
# This is also a typo. unID stickleback here should be threespine stickleback.
data.2014$fish$species_name[data.2014$fish$species_name == "unID stickleback"] <- 
  "threespine stickleback"
data.2014$fish$species_name <- droplevels(data.2014$fish$species_name)

# 2022
as.character(data.2022$fish$species_name[data.2022$fish$species_name %notin%
                              data.2022$abundance$species_name])
# No typos like this.

#### Present in abundance information but not biological ####
as.character(data.2014$abundance$species_name[data.2014$abundance$species_name %notin%
                              data.2014$fish$species_name])
# This makes sense, we cannot measure sturgeon and probably didn't have a preferred
# method of measuring periwinkle.

as.character(data.2022$abundance$species_name[data.2022$abundance$species_name %notin%
                              data.2022$fish$species_name])
# Yeah same here, we're missing some horseshoe crabs and sturgeon.
rm(date.check)

```

#### Load and clean 2018-19 data
Some data from 2018 and 2019 was not entered in the digital spreadsheets until 2023. These observations will be added to the final dataframe.

Katie- edit

```{r loading2018, results='hide', message=F, warning=F}
# 2022 data
  # Get info about all excel sheets in workbook
  fname <- paste0(here(), '/Raw_Data/Missing2018_2019_Seine_Data.xlsx')
  sheets <- readxl::excel_sheets(fname)
  # Read in as list item, each item is a sheet
  tibble <- invisible(lapply(sheets, 
                             function(x) readxl::read_excel(fname, sheet = x)))
  # Coerce list items to dataframes
  data_frame <- lapply(tibble, as.data.frame)
  # Name list items
  names(data_frame) <- sheets
  # Create temporary dataframe with all sheets
  data.2018.temp <- data_frame
  # Cut to new data list
  data.2018 <- vector(mode="list", length = 5)
  names(data.2018) <- c('sites', 'species', 'trips', 'fish', 'abundance')
  data.2018$sites <- data.2014$sites
  data.2018$species <- data.2022$species
  data.2018$trips <- data.2018.temp$trips
  data.2018$fish <- data.2018.temp$bio
  data.2018$abundance <- data.2018.temp$abund
  # Remove intermediates
  rm(data.2018.temp, data_frame, tibble, fname, sheets)

```

The species and sites dataframes do not need to be cleaned. The remainder need a quick look.

```{r siteclean2018}
#### Site Identification ####
# Sites will be the same year to year.
# What is called "Brothers" in 2022 is the same as "The Brothers - North" in 
# previous years.
```

The trip information will be cleaned similar to the 2014-2021 dataset. Freshwater sites do not exist in these years. The site numbers and loc_id variables will be merged to match the patterns that exist in the 2014-2021 dataset.

```{r tripclean2018}
#### Trip Identification #####
# Strip trip info from data list
trips <- data.2018$trips

# Remove blank rows
trips <- trips[!is.na(trips$date),]

# Convert to factor
trips$site_name <- as.factor(trips$site_name)

# Re-assign site_id
trips$site_id <- NULL
trips <- left_join(trips, dplyr::select(data.2018$sites, site_name, site_id),
                   by=c('site_name'))

# Fix substrate data type
trips$substrate <- as.factor(trips$substrate)

# Fix set time
trips$set_time <- substr(trips$set_time, start = 12, stop=19)
trips$set_time <- paste0(trips$date, " ", trips$set_time)
trips$set_time <- as.POSIXct(trips$set_time,
                               format = "%Y-%m-%d %H:%M:%S")

# Fix weather info
trips$weather <- as.factor(trips$weather)

# Fix hermit crabs 
trips$hermit_crabs[trips$hermit_crabs == 'present'] <- 1
trips$hermit_crabs[trips$hermit_crabs == 'absent'] <- 0
trips$hermit_crabs[trips$hermit_crabs == 'not circled'] <- NA
trips$hermit_crabs <- as.numeric(trips$hermit_crabs)

# Fix shrimp
trips$shrimp[trips$shrimp == 'absent'] <- 0
trips$shrimp[trips$shrimp == 'low'] <- 1
trips$shrimp[trips$shrimp == 'medium'] <- 2
trips$shrimp[trips$shrimp == 'high'] <- 3
trips$shrimp[trips$shrimp == 'very, very high'] <- 3
trips$shrimp[trips$shrimp == 'not circled'] <- NA
trips$shrimp <- as.numeric(trips$shrimp)

# Create year column
trips$year <- substr(trips$date, start=1, stop = 4)
trips$year <- as.numeric(trips$year)

# Create new trip_id based on date
#table(data.2022$trips$loc_id)
# Ended 2022 at 101, start missing data at 102
trips$new_trip_id <- as.numeric(as.factor(trips$date)) +
  max(as.numeric(as.factor(data.2018$trips$date))) + 95
#table(trips$new_trip_id)

# Create new overall information identification
trips$loc_id <- paste0(trips$year, "_",
                                 str_pad(trips$new_trip_id, 3, "left", "0"),
                                 "_",
                                 str_pad(trips$site_id, 2, "left", "0"))

# Assign bay location
trips$bay_location <- NULL
trips <- left_join(trips, bay_locations, by=c('site_name'))

# Reorder
trips <- dplyr::select(trips, loc_id, year, date, trip_id, new_trip_id, 
                       site_id, site_name, bay_location,
                       substrate, set_time, weather, temp_degc, 
                       `do_mg.l`, salinity_ppt, hermit_crabs, shrimp,
                       notes)
# QA-QC
trips$bay_location <- as.factor(trips$bay_location)
trips <- trips %>% 
  rename(`do_mg/l` = `do_mg.l`)

# Save
trips <- trips[!is.na(trips$loc_id),]
data.2018$trips <- trips
rm(trips)
```

Biological subsample information will be cleaned, but it is generally correct and doesn't need much adjustment.

```{r bioclean2018}
#### Fish Biological Information ####
bio <- data.2018$fish

# Create trip ID
bio$date <- as.POSIXct(bio$date, format='%Y-%m-%d')
pull <- dplyr::select(data.2018$trips, date, site_name, site_id, new_trip_id)
bio <- left_join(bio, pull,
                         by=c('date', 'site_name'))
bio$year <- lubridate::year(bio$date)
bio$loc_id <- paste0(bio$year, "_", bio$new_trip_id, "_", str_pad(bio$site_id, 2, 'left', '0'))

# Remove anything that doesn't have a trip_id in the trips dataframe
bio <- bio[bio$loc_id %in% data.2018$trips$loc_id,]
bio <- bio[!is.na(bio$loc_id),]

# Remove unnecessary columns
bio <- dplyr::select(bio, loc_id, species_name, sex, 
                     length_mm)

# Assign site_name and location info
bio <- left_join(bio, dplyr::select(data.2018$trips, loc_id, new_trip_id,
                                    loc_id, site_name, site_id, bay_location,
                                    date),
                 by=c('loc_id'))

# Re-order
bio <- dplyr::select(bio, loc_id, date, loc_id, new_trip_id, site_name, site_id,
                     bay_location, species_name, sex, length_mm)

# Fix data types
bio$site_name <- as.factor(bio$site_name)
bio$bay_location <- as.factor(bio$bay_location)

# Fix species names
bio$species_name[bio$species_name == "sculpin"] <- 'unID sculpin'
bio$species_name <- as.factor(bio$species_name)

# Remove instances of no catch
bio <- bio[bio$species_name %notin% c('no fish', 'no catch'),]

# Fix sex
bio$sex <- as.factor(bio$sex)

# Finalize dataframe
bio <- dplyr::select(bio,
                     loc_id, date, site_name, site_id, bay_location, 
                     species_name, sex, length_mm)
bio <- bio[!is.na(bio$species_name),]

data.2018$fish <- bio
data.2018$fish$notes <- NA
rm(bio)
```

The same is true of the abundance information.

```{r abundclean2018}
#### Abundance information ####
abund <- data.2018$abundance

# Assign loc_id
abund$loc_id <- NULL
abund <- left_join(abund, pull, by=c('date', 'site_name'))
abund$year <- lubridate::year(abund$date)
abund$loc_id <- paste0(abund$year, "_", abund$new_trip_id, "_",
                       str_pad(abund$site_id, 2, 'left', '0'))

# Remove anything that doesn't have a trip_id in the trips dataframe
abund <- abund[abund$loc_id %in% data.2018$trips$loc_id,]
abund <- abund[!is.na(abund$loc_id),]

# Assign site_name and location info
abund <- left_join(abund, dplyr::select(data.2018$trips, loc_id, bay_location),
                 by=c('loc_id'))

# Re-order
abund <- dplyr::select(abund, loc_id, date, new_trip_id, site_name, 
                       site_id, bay_location, 
                       species_name, catch)
abund <- abund[complete.cases(abund),]

# Fix data types
abund$site_name <- as.factor(abund$site_name)
abund$bay_location <- as.factor(abund$bay_location)

# Coerce non-numerics to numerics
abund$catch <- as.numeric(abund$catch)

# Fix species names
abund$species_name[abund$species_name == "sculpin"] <- 'unID sculpin'
abund$species_name <- as.factor(abund$species_name)

# Finalize
abund <- dplyr::select(abund, -new_trip_id)
data.2018$abundance <- abund
rm(abund)
```

```{r finalclean2018}
data.2018$trips$site_name <- as.factor(data.2018$trips$site_name)
data.2018$trips$bay_location <- as.factor(data.2018$trips$bay_location)
data.2018$trips <- dplyr::select(data.2018$trips, -trip_id, -new_trip_id)

test.bio <- data.2018$fish
test.bio <- dplyr::select(test.bio,
                          loc_id, date, species_name, length_mm)
test.bio <- test.bio %>%
  count(loc_id, species_name)

test.abund <- data.2018$abundance
test.abund <- dplyr::select(test.abund,
                            loc_id, date, species_name, catch)

test.combo <- left_join(test.abund, test.bio, by=c("loc_id", "species_name"))

test.bad <- test.combo[test.combo$n != test.combo$catch,]
test.bad <- test.bad[test.bad$n %notin% c(24,25, 26),]
test.bad$dif <- test.bad$n - test.bad$catch

test.bad <- test.bad[with(test.bad, order(dif)),]
test.bad <- test.bad[complete.cases(test.bad),]
test.bad <- test.bad[test.bad$dif > 0,]
# These are not typos. They are noted on the datasheets as samples missing.

# No issues with catch-bio discrepancies
rm(bay_locations, test.abund, test.bio, test.combo, i, test.bad)
```
 
As a final check, we will ensure that species names are represented correctly between biological information and abundance datasets.

```{r abundcheck2, echo=F, results='hide'}
#### Present in biological information but not in abundance ####
# 2014
as.character(data.2018$fish$species_name[data.2018$fish$species_name %notin%
                              data.2018$abundance$species_name])
# No typos.

#### Present in abundance information but not biological ####
as.character(data.2018$abundance$species_name[data.2018$abundance$species_name %notin%
                              data.2018$fish$species_name])
# no typos
rm(date.check)

```

#### Combine data
It's then a simple matter of combining all datasets to make a full record of this project.

```{r combination}
# Create blank list to collate data
data.all <- vector(mode="list", length = 5)
# Name list items
names(data.all) <- c('sites', 'species', 'trips', 'fish', 'abundance')

#### Sites ####
# Merge information from both data lists, save in combined data list
data.all$sites <- rbind(data.2014$sites, data.2022$sites)
data.all$sites <- unique(data.all$sites)

#### Species ####
data.all$species <- rbind(data.2014$species, data.2022$species)
data.all$species <- unique(data.all$species)
data.all$species <- data.all$species[data.all$species$common_name %notin%
                                       c('Add new species found in September',
                                         'Add Plankton'),]
data.all$species$ID <- NULL
data.all$species$common_name[data.all$species$common_name == "pogy"] <- 
  'atlantic menhaden'

data.all$species <- data.all$species[with(data.all$species,
                                          order(common_name)),]
rownames(data.all$species) <- NULL

#### Trips ####
data.all$trips <- rbind(data.2014$trips, data.2018$trips, data.2022$trips)
#nrow(data.all$trips)
#length(unique(data.all$trips$loc_id))
# There's one duplicate. Find and remove
#data.all$trips[data.all$trips$loc_id == "2017_042_08",]
# Can't tell why this was done. Remove both and any associated data in other tabs.
data.all$trips <- data.all$trips[data.all$trips$loc_id != "2017_042_08",]

#### Biological information ####
data.all$fish <- rbind(data.2014$fish, data.2018$fish, data.2022$fish)
data.all$fish <- data.all$fish[data.all$fish$loc_id != "2017_042_08",]

#### Abundance ####
data.all$abundance <- rbind(data.2014$abundance, data.2018$abundance, data.2022$abundance)
data.all$abundance <- data.all$abundance[data.all$abundance$loc_id != "2017_042_08",]

```

## Save data
Combined data can be saved as a new master seine spreadsheet.

```{r savedata, eval=F}
library(xlsx)
write.xlsx2(data.all$sites, 
            here('Clean_Data/compiled_clean.xlsx'),
            sheetName = 'sites',
            append=FALSE,
            row.names = FALSE)
write.xlsx2(data.all$species, 
            here('Clean_Data/compiled_clean.xlsx'),
            sheetName = 'species',
            append=TRUE,
            row.names = FALSE)
write.xlsx2(data.all$trips, 
            here('Clean_Data/compiled_clean.xlsx'),
            sheetName = 'trips',
            append=TRUE,
            row.names = FALSE)
write.xlsx2(data.all$fish, 
            here('Clean_Data/compiled_clean.xlsx'),
            sheetName = 'bio',
            append=TRUE,
            row.names = FALSE)
write.xlsx2(data.all$abundance, 
            here('Clean_Data/compiled_clean.xlsx'),
            sheetName = 'abund',
            append=TRUE,
            row.names = FALSE)

```
